%%     16 preface
 % \chapter{Preface}
 % 
 % Some people complain that Computer Science is not a science, and
 % that the fact is pointed up by the presence of the word ``science''
 % in the title.  I'm not sure whether these people want to change the
 % name to emphasize that it is not a science, or to change it in
 % order to cause it to become a science, or if they somehow want to
 % change it into a science by some other means.  I, for one, revel in
 % the non-science status of the field.  Kuhn tells us that all fields
 % go through a pre-scientific stage, and describes the groundwork
 % which is laid during that period for the science that is to follow.
 % It is clear, however, that the idea of ``doing science'' in a field
 % before it has {\em become} a science is an absurdity.
 %
 % Thus, I submit the following ``system thesis.''
%%     34 abstract

\renewcommand{\baselinestretch}{1.0}\tiny\normalsize
\begin{tabbing}
{\em Advisor:} \= \kill
{\em Title:} \> Tabula Rasa: A Multi-scale User Interface System \\
{\em Author:} \> David Fox \\
{\em Advisor:} \> Ken Perlin
\end{tabbing}
\renewcommand{\baselinestretch}{\spacing}\tiny\normalsize

%\begin{center}
%{\bf Dissertation Abstract}
%\end{center}

This dissertation develops the concept of a zoomable user interface and
identifies the design elements which are important to its viability as
a successor to the desktop style of interface.  The implementation of
an example system named {\em Tabula Rasa} is described, along with the
design and implementation of some sample applications for Tabula Rasa.
We show how programming techniques such as delegation and
multi-methods can be used to solve certain problems that arise in
the implementation of Tabula Rasa, and in the implementation of Tabula
Rasa applications.

Over the past thirty years the desktop or {\sc WIMP} (Windows, Icons,
Menus, Pointer) user interface has made the computer into a tool that
allows non-specialists to get a variety of tasks done.  In recent
years, however, the applications available under this interface have
become larger and more unwieldy, taking into themselves more and more
marginally related functionality.  Any inter-operability between
applications must be explicitly designed in.

The Zoomable User Interface (ZUI) is a relatively new metaphor designed as
a successor to the desktop interface.  It is inspired by the Pad system,
which is based on a zoomable surface of unlimited resolution.  Just as
the desktop interface has a set of essential elements, a ZUI has a set
of elements each of which is vital to the whole.  These include
\begin{list}{$\bullet$}{\itemsep 0pt}
\item a zoomable imaging model,
\item a persistent virtual geography for data objects,
\item {\em semantic zooming} to optimize the utility of screen space,
\item {\em work-through interfaces} for application objects,
\item a constraint system for ensuring the consistency of the
        interface elements.
\end{list}
These basic elements combine to produce an environment that takes
advantage of the user's spatial memory to create a more expansive
and dynamic working environment, as well as encouraging finer grained
applications that automatically inter-operate with various types of
data objects and applications.

\pagebreak

%%      4 Contents
